
\documentclass{article}

\begin{document}

\title{Proxy objects in La-Tng.}

\author{I. P. McCulloch}
\date{\today}

\maketitle



C++ doesn't support transparent proxies.  Expression types in the
matrix library don't own the data they refer to, and they act somewhat
like C++ reference types.  However, unlike general-purpose proxies,
these proxy types fully satisfy the corresponding interface requirements.

Proxies follow the logical const'ness of the object they are proxy
for.  This is different to the way C++ references work, in in that a
const reference \verb"T& const" is not a type distinct from
'verb"T\&". The former can only be constructed in a template or typedef
and the const is simply dropped. 

This means that \verb"proxy<T> const" and \verb"proxy<T const>" are
broadly equivalent types. Indeed, there should be a constructor for 
\verb"proxy<T const>" that accepts a \verb"proxy<T> const".

Ideally, the copy constructor for a non-const proxy should require a non-const
reference. If the copy ctor passed a non-const reference, it violates
const-correctness because it is then possible to construct a non-const
proxy from a const proxy. But because of language rules forbidding
binding a temporary to a non-const reference, this solution is not
practical at this time. Hence we must accept a (limited) violation of
const correctness. Note that this only applies to \verb"proxy<T> const"; 
if you instead have a \verb"proxy<T const>" you cannot
violate const correctness.

Functor types:

non-proxies:

\verb"Func<T>" should have an argument\_type that is either
\verb"T" or \verb"T const&".

\verb"Func<proxy>" should have an argument\_type that is either 
\verb"T const" or \verb"T const&". Note that this convention also
allows for value types, if one desires the formal parameter to be
always const.

\verb"Func<T&>" should have an argument\_type of \verb"T&", for both
proxies and values.

Rule of thumb: given template argument \verb"T", the argument\_type
will be \verb"T const".

An \textit{adaptor} is a value type that contains another value type.
For example, a sparse vector that uses \verb"std::map" as its
underlying container. This is different from a proxy in that the
adaptor controls the lifetime of the data it contains, whereas a proxy
does not. In practice, one template class can take on both roles,
depending on whether the template is a proxy-reference or a value.

If \verb"Container<T>" stores an object of type \verb"T", then T can
be either a const or non-const proxy, const or non-const reference, or
a (non-const) value. Whether the container is a proxy or an adaptor is
decided by \verb"is_proxy_reference<T>".

The basic idea behind \verb"make_reference" and \verb"make_const_reference"
is that when constructing proxy classes themselves, you want to store
value types by reference, but proxy types by the proxy itself (NOT a
reference to the proxy). Indeed, you \textit{must} do this, as the 
proxy is likely to be a temporary.

MakeReference:
\verb"T" $\rightarrow$ \verb"T\&"
\verb"T const" $\rightarrow$ \verb"T const\&"
\verb"T\&" $\rightarrow$ \verb"T\&"
\verb"T const\&" $\rightarrow$ \verb"T const\&"
\verb"proxy" $\rightarrow$ \verb"proxy"
\verb"proxy const" $\rightarrow$ \verb"proxy const"
\verb"proxy\&" $\rightarrow$ \verb"proxy"
\verb"proxy const\&" $\rightarrow$ \verb"proxy const"

MakeConstReference:
\begin{tabular}{l}
\verb"T" $\rightarrow$ \verb"T const\&"                  \\
\verb"T const" $\rightarrow$ \verb"T const\&"  \\
\verb"T\&" $\rightarrow$ \verb"T const\&"  \\
\verb"T const\&" $\rightarrow$ \verb"T const\&"  \\
\verb"proxy" $\rightarrow$ \verb"proxy const"  \\
\verb"proxy const" $\rightarrow$ \verb"proxy const"  \\
\verb"proxy\&" $\rightarrow$ \verb"proxy const"  \\
\verb"proxy const\&" $\rightarrow$ \verb"proxy const"
\end{tabular}

ArgType:
\verb"T" $\rightarrow$ \verb"T"
\verb"T const" $\rightarrow$ \verb"T" 
\verb"T\&" $\rightarrow$ \verb"T\&" 
\verb"T const\&" $\rightarrow$ \verb"T" 
\verb"proxy" $\rightarrow$ \verb"proxy" 
\verb"proxy const" $\rightarrow$ \verb"proxy" 
\verb"proxy\&" $\rightarrow$ \verb"proxy\&" 
\verb"proxy const\&" $\rightarrow$ \verb"proxy" 


\end{document}